Reinforcement Learning Produces Dominant Strategies for the Iterated Prisoner's Dilemma

### 2017-07-11 main.tex
Various Minor Edits

The iterated prisoner's dilemma (IPD) is an area of game theoretic research
aiming to understand the evolution of cooperative behaviour from complex
dynamics \cite{Axelrod1984}.

The Axelrod library \cite{axelrodproject, knight2016open} is an open source
software for conducting IPD research with
reproducibility as a principal goal. Written in the Python programming language,
to date the library contains source code contributed by over 50 individuals
from a variety of geographic locations and technical backgrounds. The library
is supported by a comprehensive test suite that covers all the intended
behaviors of the strategies in the library, as well as the features that
conduct matches, tournaments, and population dynamics.

@@

Since Robert Axelrod's seminal tournament \cite{Axelrod1980a}, a number of
IPD\@ tournaments have been undertaken and are summarised in
Table~\ref{tbl:tournaments}. Further to the work described in
\cite{knight2016open} a regular set of standard, noisy and probabilistic ending
tournaments are carried out as more strategies are added to the Axelrod library.
Details and results are available here:
\url{http://axelrod-tournament.readthedocs.io}. This work presents a detailed
analysis of a tournament with 176 strategies, more details about this are given
in Section~\ref{sec:results}.

@@

In this work we describe how collections of strategies in the Axelrod library
have been used to train new strategies specifically to win IPD tournaments.
These strategies are trained using generic strategy archetypes based on e.g.\
finite state
machines, arriving at particularly effective parameter choices through
evolutionary or particle swarm algorithms. There are several
previous publications that use evolutionary algorithms to
evolve IPD strategies in various circumstances
\cite{ashlock2006training, Ashlock2015, Ashlock2006,
      ashlock2014shaped, Ashlock2014, barlow2015varying,
      fogel1993evolving, marks1989niche, sudo2015effects,
      vassiliades2010multiagent}. See also \cite{Gaudesi2016} for a
strategy trained to win against a collection of well-known IPD opponents and see
\cite{franken2005particle} for a prior use of particle swarm algorithms. Our
results are unique in that we are able to train against a large collection of
well-known strategies available in the scientific literature. Crucially, the
software used in this work is openly available and can be used to train strategies
in the future in a reliable manner, with confidence that the opponent strategies
are correctly implemented, tested and documented.
Moreover, as of the time of writing, we claim that this work contains the best
performing strategies for the iterated prisoner's dilemma.

@@

Training of this strategy corresponds to finding maps from partial histories to
either a cooperation or a defection. Although various
combinations of $n_1, m_1,$ and $m_2$ have been tried, the best performance at
the time of
training was obtained for $n_1 = m_1 = m_2 = 2$ and generally for $n_1 > 0$.
A strategy
called EvolvedLookerUp2\_2\_2 is among the top strategies in the library.

This archetype can be used to train deterministic memory-$n$ strategies with the
parameters $n_1=0$ and $m_1=m_2=n$. For $n=1$, the resulting strategy cooperates
if the last round was mutual cooperation and defects otherwise.

@@

Training of this strategy corresponds to finding maps from histories to
a probability of cooperation. The library includes a strategy trained
with $n_1 = m_1 = m_2 = 2$ that is \emph{mostly deterministic}, with most of the
probabilities being 0 or 1. At one time this strategy outperformed
EvolvedLookerUp2\_2\_2.

This strategy type can be used to train arbitrary memory-$n$ strategies. A
memory one strategy called PSOGamblerMem1 was trained, with
probabilities $(\text{Pr}(\text{C}\;|\;\text{CC}),
                \text{Pr}(\text{C}\;|\;\text{CD}),
                \text{Pr}(\text{C}\;|\;\text{DC}),
                \text{Pr}(\text{C}\;|\;\text{DD})) = (1, 0.5217, 0, 0.121)$.
Though it performs well in standard tournaments (see
Table~\ref{tbl:standard_score})
it is not as good as the longer memory strategies, and is bested by a similar
strategy that also uses the first round of play: PSOGambler\_1\_1\_1.

@@

Because these strategies inherit many of the properties of the strategies
on which they are based, including using the match length to defect on the last
round(s) of play, not all of these strategies were included in results of this
paper. These strategies do not typically outperform the trained strategies
described above.

@@

Finally, Table~\ref{tbl:standard_ranks} and
Figure~\ref{fig:standard_ranks_boxplot} show the ranks (based on median score)
of each strategy over the repeated tournaments. Whilst there is some
stochasticity, the top three strategies almost always rank in the top three. For
example, the worst that the Evolved Lookerup 2 2 2 ranks in a given tournament
is 8th.

@@

Noisy tournaments in which there is a 5\% chance that an action is
flipped are now described. As shown in Table~\ref{tbl:noisy_score} and
Figure~\ref{fig:noisy_score}, the best performing strategies in median payoff
are DBS, designed to correct for noise, followed by two strategies trained in
the presence of noise and three trained strategies trained without noise. One of
the strategies trained with noise (PSO Gambler) actually performs less well than
some of the other high ranking strategies including
Spiteful TFT (TFT but defects indefinitely if the opponent defects twice
consecutively) and OmegaTFT (also designed to handle noise).

@@

As shown in Table~\ref{tbl:noisy_wins_top_winners} and
Figure~\ref{fig:noisy_winplot} the strategies tallying the most wins are
somewhat similar to the Standard tournaments, with Defector, the handshaking
CollectiveStrategy, and Aggravate appearing as the top three again.

@@

Finally, Table~\ref{tbl:noisy_ranks} and
Figure~\ref{fig:noisy_ranks_boxplot} show the ranks (based on median score)
of each strategy over the repeated tournaments. We see that the stochasticity
of the ranks understandably increases relative to the Standard tournament. An
exception is the top three strategies, for example, the DBS strategy never ranks
lower than
second and wins 75\% of the time. The two strategies trained for noisy
tournaments rank in the top three 95\% of the time.
% TODO Possibly update this with more data.

@@

Figure~\ref{fig:comparison_cooperation_heatmaps_noisy} shows the rate of
cooperation in each round for the top three strategies and just as for the top
performing strategies in the standard tournament
(Figure~\ref{fig:comparison_cooperation_heatmaps_standard}) it is evident that
the strategies never defect first and learn to quickly punish poorer strategies.

@@

The Particle Swarm Algorithm is implemented using the Python library:
\url{https://pypi.python.org/pypi/pyswarm}. This algorithm was used only to
train the Gambler archetype.

@@

These can be used in Noisy or Standard environments (as evidenced by
Sections~\ref{sec:standard} and~\ref{sec:noisy}). These objectives can be
further modified to suit other purposes. New strategies could be trained with
variations including spatial structure, and probabilistically ending matches.

@@

The tournament results indicate that pre-trained strategies are generally better
than human designed strategies at maximizing payoff against a diverse set of
opponents. An evolutionary algorithm produces strategies based on multiple
generic archetypes that are able to achieve a higher average
score than any other known opponent in a standard tournament. Most of the trained
strategies use multiple rounds of the history of play (some using all of it) and
outperform memory-one strategies from the literature. Interestingly, a trained
memory one strategy produced by a particle swarm algorithm performs well, better
than human designed strategies such as Win Stay Lose Shift and zero determinant
strategies (which enforce a payoff difference rather than maximize total payoff).
The generic structure of the trained strategies did not appear to be
critical for the standard tournament -- strategies based on lookup tables,
finite state machines, neural networks, and stochastic variants all performed well.
Single layer neural networks (Section~\ref{sec:ann}) performed well in both
noisy and standard tournaments
though these had some aspect of human involvement in the selection of features.
This is in line with the other strategies also where some human decisions are made
regarding the structure. For the LookerUp and Gambler archetypes
(Sections~\ref{sec:lookerup} and~\ref{sec:gambler})
a decision has to be made
regarding the number of turns of history and initial play that are to be used.
In contrast, the finite state machines and hidden Markov models
(Sections\ref{sec:fsm} and~\ref{sec:hmm}) required only a choice of the number
of states, and the training algorithm can eliminate unneeded states in the case
of finite state machines (evidenced by the unconnected nodes in the diagrams
for the included representations).

Many strategies can be represented by multiple archetypes, however some
archetypes will be more efficient in encoding the patterns present in the data.
The fact that the Lookerup strategy does the best for the standard
tournament indicates that it represents an efficient reduction of
dimension which in turn makes its training more efficient. In particular the
first rounds of play were valuable bit of information. For the noisy
tournament however the dimension reduction represented by some archetypes
indicates that some features of the data are not captured by the lookup
tables while they are by the neural network and the finite state machine,
allowing them to adapt better to the noisy environment. Intuitively, a noisy
environment can significantly affect a lookup table based on the last two rounds
of play since these action pairs compete with probing defections, apologies, and
retaliations. Accordingly, it is not surprising that additional parameter space
is needed to adapt to a noisy environment.

@@

Finally, we note that as the library grows, the top performing strategies
sometimes shuffle, and are not retrained automatically. Most of the strategies were
trained on an earlier version of the library (v2.2.0: \cite{axelrodproject2.2})
that did not include DBS and several other opponents. The precise parameters
that are optimal will depend on the pool of opponents. Moreover we have not
extensively trained strategies to determine the minimum parameters that are
sufficient -- neural networks with fewer nodes and features and finite state
machines with fewer states may suffice. See \cite{ashlock2013impact} for
discussion of resource availability for IPD strategies.

>>>>

The iterated prisoner's dilemma (IPD) is a common model in game theory,
frequently used  to understand the evolution of cooperative behaviour from complex
dynamics \cite{Axelrod1984}.

In this manuscript we use the Axelrod library \cite{axelrodproject, knight2016open},
open source software for conducting IPD research with
reproducibility as a principal goal. Written in the Python programming language,
to date the library contains source code contributed by over 50 individuals
from a variety of geographic locations and technical backgrounds. The library
is supported by a comprehensive test suite that covers all the intended
behaviors of all of the strategies in the library, as well as the features that
conduct matches, tournaments, and population dynamics.

@@

Since Robert Axelrod's seminal tournament \cite{Axelrod1980a}, a number of
IPD\@ tournaments have been undertaken and are summarised in
Table~\ref{tbl:tournaments}. Further to the work described in
\cite{knight2016open} a regular set of standard, noisy and probabilistic ending
tournaments are carried out as more strategies are added to the Axelrod library.
Details and results are available here:
\url{http://axelrod-tournament.readthedocs.io}. This work presents a detailed
analysis of a tournament with 176 strategies (details given
in Section~\ref{sec:results}).

@@

In this work we describe how collections of strategies in the Axelrod library
have been used to train new strategies specifically to win IPD tournaments.
These strategies are trained using generic strategy archetypes based on e.g.\
finite state
machines, arriving at particularly effective parameter choices through
evolutionary or particle swarm algorithms. There are several
previous publications that use evolutionary algorithms to
evolve IPD strategies in various circumstances
\cite{ashlock2006training, Ashlock2015, Ashlock2006,
      ashlock2014shaped, Ashlock2014, barlow2015varying,
      fogel1993evolving, marks1989niche, sudo2015effects,
      vassiliades2010multiagent}. See also \cite{Gaudesi2016} for a
strategy trained to win against a collection of well-known IPD opponents and see
\cite{franken2005particle} for a prior use of particle swarm algorithms. Our
results are unique in that we are able to train against a large and diverse
collection of strategies available from the scientific literature.
Crucially, the
software used in this work is openly available and can be used to train strategies
in the future in a reliable manner, with confidence that the opponent strategies
are correctly implemented, tested and documented.
Moreover, as of the time of writing, we claim that this work contains the best
performing strategies for the iterated prisoner's dilemma.

@@

Training of this strategy corresponds to finding maps from partial histories to
actions, either a cooperation or a defection. Although various
combinations of $n_1, m_1,$ and $m_2$ have been tried, the best performance at
the time of
training was obtained for $n_1 = m_1 = m_2 = 2$ and generally for $n_1 > 0$.
A strategy
called EvolvedLookerUp2\_2\_2 is among the top strategies in the library.

This archetype can be used to train deterministic memory-$n$ strategies with the
parameters $n_1=0$ and $m_1=m_2=n$. For $n=1$, the resulting strategy cooperates
if the last round was mutual cooperation and defects otherwise, known as Grim or
Grudger.

@@

Training of this strategy corresponds to finding maps from histories to
a probability of cooperation. The library includes a strategy trained
with $n_1 = m_1 = m_2 = 2$ that is \emph{mostly deterministic}, with 12 of the 64
probabilities being 0 or 1. At one time this strategy outperformed
EvolvedLookerUp2\_2\_2.

This strategy type can be used to train arbitrary memory-$n$ strategies. A
memory one strategy called PSOGamblerMem1 was trained, with
probabilities $(\text{Pr}(\text{C}\;|\;\text{CC}),
                \text{Pr}(\text{C}\;|\;\text{CD}),
                \text{Pr}(\text{C}\;|\;\text{DC}),
                \text{Pr}(\text{C}\;|\;\text{DD})) = (1, 0.5217, 0, 0.121)$.
Though it performs well in standard tournaments (see
Table~\ref{tbl:standard_score})
it does not outperform the longer memory strategies, and is bested by a similar
strategy that also uses the first round of play: PSOGambler\_1\_1\_1.

@@

Because these strategies inherit many of the properties of the strategies
on which they are based, including using knowledge of the match length to defect
on the last round(s) of play, not all of these
strategies were included in results of this
paper. These strategies do not typically outperform the trained strategies
described above.

@@

Finally, Table~\ref{tbl:standard_ranks} and
Figure~\ref{fig:standard_ranks_boxplot} show the ranks (based on median score)
of each strategy over the repeated tournaments. Whilst there is some
stochasticity, the top three strategies almost always rank in the top three. For
example, the worst that the Evolved Lookerup 2 2 2 ranks in any tournament
is 8th.

@@

Noisy tournaments in which there is a 5\% chance that an action is
flipped are now described. As shown in Table~\ref{tbl:noisy_score} and
Figure~\ref{fig:noisy_score}, the best performing strategies in median payoff
are DBS, designed to account for noise, followed by two strategies trained in
the presence of noise and three trained strategies trained without noise. One of
the strategies trained with noise (PSO Gambler) actually performs less well than
some of the other high ranking strategies including
Spiteful TFT (TFT but defects indefinitely if the opponent defects twice
consecutively) and OmegaTFT (also designed to handle noise).

@@

As shown in Table~\ref{tbl:noisy_wins_top_winners} and
Figure~\ref{fig:noisy_winplot} the strategies tallying the most wins are
somewhat similar to the standard tournaments, with Defector, the handshaking
CollectiveStrategy, and Aggravate appearing as the top three again.

@@

Finally, Table~\ref{tbl:noisy_ranks} and
Figure~\ref{fig:noisy_ranks_boxplot} show the ranks (based on median score)
of each strategy over the repeated tournaments. We see that the stochasticity
of the ranks understandably increases relative to the standard tournament. An
exception is the top three strategies, for example, the DBS strategy never ranks
lower than
second and wins 75\% of the time. The two strategies trained for noisy
tournaments rank in the top three 95\% of the time.
% TODO Possibly update this with more data.

@@

Figure~\ref{fig:comparison_cooperation_heatmaps_noisy} shows the rate of
cooperation in each round for the top three strategies (in the absense of noise)
and just as for the top performing strategies in the standard tournament
(Figure~\ref{fig:comparison_cooperation_heatmaps_standard}) it is evident that
the strategies never defect first and learn to quickly punish poorer strategies.

@@

The Particle Swarm Algorithm is implemented using the pyswarm library:
\url{https://pypi.python.org/pypi/pyswarm}. This algorithm was used only to
train the Gambler archetype.

@@

These can be used in Noisy or Standard environments (as evidenced by
Sections~\ref{sec:standard} and~\ref{sec:noisy}). These objectives can be
further modified to suit other purposes. New strategies could be trained with
variations including spatial structure and probabilistically ending matches.

@@

The tournament results indicate that pre-trained strategies are generally better
than human designed strategies at maximizing payoff against a diverse set of
opponents. An evolutionary algorithm produces strategies based on multiple
generic archetypes that are able to achieve a higher average
score than any other known opponent in a standard tournament. Most of the trained
strategies use multiple rounds of the history of play (some using all of it) and
outperform memory-one strategies from the literature. Interestingly, a trained
memory one strategy produced by a particle swarm algorithm performs well, better
than human designed strategies such as Win Stay Lose Shift and zero determinant
strategies (which enforce a payoff difference rather than maximize total payoff).
The generic structure of the trained strategies did not appear to be
critical for the standard tournament -- strategies based on lookup tables,
finite state machines, neural networks, and stochastic variants all performed well.
Single layer neural networks (Section~\ref{sec:ann}) performed well in both
noisy and standard tournaments
though these had some aspect of human involvement in the selection of features.
This is in line with the other strategies also where some human decisions are made
regarding the structure. For the LookerUp and Gambler archetypes
(Sections~\ref{sec:lookerup} and~\ref{sec:gambler})
a decision has to be made
regarding the number of rounds of history and initial play that are to be used.
In contrast, the finite state machines and hidden Markov models
(Sections\ref{sec:fsm} and~\ref{sec:hmm}) required only a choice of the number
of states, and the training algorithm can eliminate unneeded states in the case
of finite state machines (evidenced by the unconnected nodes in the diagrams
for the included representations).

Many strategies can be represented by multiple archetypes, however some
archetypes will be more efficient in encoding the patterns present in the data.
The fact that the Lookerup strategy does the best for the standard
tournament indicates that it represents an efficient reduction of
dimension which in turn makes its training more efficient. In particular the
first rounds of play were valuable bits of information. For the noisy
tournament however the dimension reduction represented by some archetypes
indicates that some features of the data are not captured by the lookup
tables while they are by the neural networks and the finite state machines,
allowing the latter to adapt better to the noisy environment. Intuitively, a noisy
environment can significantly affect a lookup table based on the last two rounds
of play since these action pairs compete with probing defections, apologies, and
retaliations. Accordingly, it is not surprising that additional parameter space
is needed to adapt to a noisy environment.

@@

Finally, we note that as the library grows, the top performing strategies
sometimes shuffle, and are not retrained automatically. Most of the strategies were
trained on an earlier version of the library (v2.2.0: \cite{axelrodproject2.2})
that did not include DBS and several other opponents. The precise parameters
that are optimal will depend on the pool of opponents. Moreover we have not
extensively trained strategies to determine the minimum parameter spaces that are
sufficient -- neural networks with fewer nodes and features and finite state
machines with fewer states may suffice. See \cite{ashlock2013impact} for
discussion of resource availability for IPD strategies.
